Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2023-12-06 demystify.tex
updates

Most CS PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can greatly strengthen your application. However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, mentors) that know you well enough to talk about \emph{your research experience and capabilities}. This person can be a postdoc who has mentored you.

\begin{commentbox}[Vu:]
    If you are from a good school from your country or your grading system is not standard. You can ask your reference writers to explain about that in their letters.  For example, "Bac Khoa" are the top universities in Vietnam for CS but few people outside Vietnam know about them.  So if you are from there, you should ask your reference writers to mention that in their letters.
\end{commentbox}

Many students have letters written by themselves and signed by their professors. Such letters have little
values and would be considered weak (you cannot even find someone cares enough about you to write a candid personal reference letter?).  Instead of the reference writer writing about you, in this case it is you who write about yourself (and they just sign the letter).

Also, many professors write generic letters for students, e.g., the students didn't do any
research or make any impression for the professor to write about. These letters are also not useful and considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well ask me to write letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their profile.

Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS).

@@

\paragraph{Always Waive your Right to Review the LORs}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers, and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{Reminding ref. writers to submit their letters} After you submit your application, you can tell your writers that you have submitted your application and remind them to submit their letters if they haven't done so. But don't send too many reminders as that can be annoying to the writers.

Note that most places only have deadlines for you, the applicant, but are very flexible with the letter writers (in many cases do not really give them any deadline or a very general one).  Also, many places do not begin the admission review process right after the deadline and instead won't start until the next semester (mid January).

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

@@

\emph{MS Degree requirement}:  as mentioned in \S\ref{sec:msrequirement} and \ref{sec:time}, US PhD programs do not require MS degree.  In contrast, many other countries do require MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

@@

\emph{Course work}: in the US you will spend the first couple of years taking classes and explore potential adviser and research topics. In other countries, you often do not do course work and start your research right away, e.g., you immediately work on the research project you proposed with the adviser you chose. Also, in the US you also have to pass a series of "exams", e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word "ABD" (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}. In other countries, you do not have to do these exams or only do few of them.

\emph{Funding}:  In many countries stipend comes from the university or from gov't research organizations and often have a fixed duration (e.g., 3 years).  In the US, stipend (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in the European universities compared to the US.

@@

\emph{Work-life balance}: I believe that PhD students in the US are overworked compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to do TA in addition to their own research. In contrast, PhD students in other countries are often paid through fellowships, which do not require them to do TA.

@@

\begin{commentbox}[DK:]
  I was amused to read this as if I recall correctly, you never called me by my first name when you were at UNM. You always called me Prof. And, many times, I would jokingly call you back as Prof. Vu.
  \tcblower
  \textbf{Vu}: Yes, for some reason I enjoy addressing you as ``Prof.'' (without appending a last or first name).  I call Steph and everyone else by their their first names.  The use of Prof. Vu may have foreshadowed my future in academia.
\end{commentbox}

Certainly, there's a peculiar satisfaction in addressing you as "Prof." without appending a last or first name. While I refer to Steph and everyone else by their first names, perhaps addressing Prof. Vu in this manner hinted at my future in academia.

>>>>

Most PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can greatly strengthen your application. However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS).

\begin{commentbox}[Vu:]
    If your grading system is not US standard or you are from a good school unknown outside of your country, you can ask your reference writers to explain about that in their letters.  For example, "Bach Khoa" are the top universities in Vietnam for STEM studies but few people outside Vietnam know about them.  So if you are from there, you should ask your reference writers to mention that.
\end{commentbox}

\paragraph{Self-written Letters} Many international students write letters themselves (typically due to the requests of their profs.) and have them signed by their profs. Such letters have little values and would be considered weak by reviewers, who can question why you cannot even find someone cares enough about you to write a candid personal reference letter.  Instead of the reference writer writing about you, in this case it is you who write about yourself (and they just sign the letter).

\paragraph{Generic Letters} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic letter, which is not useful and also considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well ask me to write letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

\paragraph{Waiving your Right to Review LORs}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers, and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{Reminding Writers to Submit Letters} After you submit your application, you can tell your writers that and remind them to submit their letters if they haven't done so. But don't send too many reminders as that can be annoying to the writers.

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid January).

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries, which might require MS first (recall that CS PhD programs in the US do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

@@

\emph{MS Degree requirement}:  as mentioned in \S\ref{sec:msrequirement} and \ref{sec:time}, CS PhD programs in the US do not require MS degree.  In contrast, many other countries do require MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

@@

\emph{Course work}: in the US you will spend the first couple of years taking classes and explore potential adviser and research topics. In other countries, you (who already have an MS) start your research right away, e.g., you immediately work on the research project you proposed with the adviser you chose. Also, in the US you also have to pass a series of "exams", e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word "ABD" (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}. In other countries, you do not have to do these exams or only do a few of them.

\emph{Funding}:  In many countries stipend comes from the university or from the gov't. These funding often have a fixed duration (e.g., 3 years).  In the US, stipend (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in the European universities compared to the US.

@@

\emph{Work-life balance}: PhD students in the US are often said to be overworked compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to do TA in addition to their own research. In contrast, PhD students in other countries are often paid through fellowships, which do not require them to do TA.

@@

\begin{commentbox}[DK:]
  I was amused to read this as if I recall correctly, you never called me by my first name when you were at UNM. You always called me Prof. And, many times, I would jokingly call you back as Prof. Vu.
  \tcblower
  \textbf{Vu}: Yes, for some reason I enjoy addressing you as ``Prof.'' (without appending a last or first name).  The use of Prof. Vu may have foreshadowed my future in academia.
\end{commentbox}
